Mapping Epigraphic Databases to EpiDoc

### 2016-01-20 source/17/Zuddas Epigraphy in Italian High Schools (final).tex
fixed few more papers
minor changes. THERE HAS BEEN CONFUSION WITH THE BIBLIO, MASTER AND
SOME OTHER FILES

The use of local materials, which can be checked out personally, may more easily arouse the learners’ attention. That is
why in this paper I will mainly refer to inscriptions from modern Umbria (regiones VI and VII, from Perusia) and
especially from Asisium, used at school during my experimental lessons. The local museum, best known as the ``Foro
romano'', hosts a huge collection of inscriptions that were catalogued in 2008 by pupils of the Liceo Properzio, under
my supervision, with details of the type of inscription, material, place of origin and chronology. Previously a booklet
about Roman Assisi had been
written\footnote{\url{http://www.liceoassisi.it/index.php?option=com\_content\&view=article\&id=47:assisi-romana\&catid=6:i-nostri-risultati\&Itemid=34.}},
both in Italian and English, comprising a page about the famous tetrastyle with a translation of the inscription on the
base (CIL 11, 05372 = EDR025323, later inserted in the MediaWiki page), which recalls the official inauguration of the
aedicule: at the time, as was the Roman custom, money was given to decurions, seviri Augustales and the common people.
Thanks to these few lines, the class is taken on a journey to ancient Asisium, and gets a lot of feedback about
religion, architecture, social life and economy in the first century A.D.

@@

\begin{itemize}
\item elements of Latin grammar, syntax and linguistics\footnote{\citet{Hartnett2012}; \citet{McCarthy1992}.};
\item culture and history;
\item and also, to a lesser extent, lexicon. 
\end{itemize}

@@

\begin{itemize}
\item the ones that are relevant for their Italian derivates (e.g. os, oris);
\item the ones that belong to the most significant semantic fields (always keeping in mind that some words are more
important than others: for instance, knowing the difference between bellum and pugna helps to develop the comprehension
of the two different categories; on the contrary, a non-specialized student does not need to know who a primus pilus
was);
\item the abstract ones that are fundamental to understanding the Roman way of thinking (e.g. imperium, virtus, fas).
\end{itemize}
Accordingly, a teacher should concentrate on those aspects of Roman society that are still pre-eminent for us:
archaeology (Roman buildings, domus, theatres, roads), myths and religion, politics (Empire, war and globalization).
Inscriptions certainly offer a less varied lexical repertoire than a literary text, but can still help to give
substance to these contexts.

@@

\begin{quotation}
P. Decimius P. l. Eros / Merula, medicus / clinicus, chirurgus, / ocularius, VIvir. / Hic pro libertate dedit
((sestertium)) (quinquaginta milia). / Hic pro seviratu in rem p(ublicam) / dedit ((sestertium)) (duo milia). / Hic in
statuas ponendas in / aedem Herculis dedit ((sestertium)) (triginta milia). / Hic in vias sternendas in / publicum
dedit ((sestertium)) (triginta septem milia). / Hic pridie quam mortuus est / reliquit patrimoni ((sestertium))... 
\end{quotation}

not only does a student make contact with the technical lexicon of medicine and words related to building activities
(statuas ponere, vias sternere), but he can also perceive the ``evergetic spirit'' of an ancient society: a physician was
usually a freedman coming from the East, who could be very rich and spend his money for public utility; whenever a
person obtained a priesthood, he used to pay a summa honoraria. The text can also be used for linguistic purposes
(prepositions like pro; partitive genitive; deponent verbs).

This does not mean that inscriptions are not useful for seeing specific abstract words in their context: when talking
about Roman virtues and mos maiorum, the famous Clipeus from Arles (\emph{AE} 1952, 0165 = \emph{AE} 1994, 0227) is a
perfect complement to chapter 34 of the Res Gestae (formally an inscription, too): \begin{quotation}
\emph{Senatus / populusque Romanus /
imp(eratori) Caesari / Divi f(ilio) / Augusto / co(n)s(uli) VIII dedit clupeum / virtutis, clementiae, / iustitiae,
pietatis erga / deos patriamque.}

@@

For instance, if I say Iulia rosas amat, even an Italian who knows no Latin can get the meaning of the sentence.
Problems are encountered when the sentences and texts to translate become more complex: only when it is too late does
an ill-prepared teacher realize that the class has not internalized the language system and the patterns.

In funerary inscriptions the value of cases is essential. The texts can be simple, but if you want to understand who is
dead, who is the dedicant and what the relationship between them is, you have to distinguish dative and nominative
cases. Moreover, the concordance of the different elements of the onomastics (especially praenomina and filiation) is
good way to encourage (and, for a teacher, to test) the learning of declensions not in a merely mnemonic, but also in
an active way.

@@

\begin{itemize}
\item CIL 11, 05501 = EDR025449: \begin{quotation}
Noniae Privatae, C. Propertius -{}-{}-{}-{}-{}-
\end{quotation}

@@

\item CIL 11, 05461 = EDR025411: \begin{quotation}
A(n)noru(m) XIX. Calventia C(ai) f(ilia) Polla, L(ucius)
Vistinius vir, Gavia mater posuer(unt).

@@

\item CIL 11, 05399 = EDR025349: \begin{quotation}
P(ublius) Decimius P(ubli) l(ibertus) Eros Merula VIvir
viam a cisterna ad domum L(uci) Muti stravit ea pecunia -{}-{}-{}-{}-{}- 
\end{quotation}

With an inscription like this you can address different topics, both historical (the importance of freedmen, the
imperial cult) and linguistic ones (indirect complements, uses of is ea id as an adjective and to introduce a relative
clause that, in this case, has evidently been lost).
\end{itemize}

@@

\begin{quotation}
Post(umus) Mimesius C(ai) f(ilius), T(itus) Mimesius Sert(oris) f(ilius), Ner(o) Capidas C(ai) f(ilius) Ruf(- - -),
Ner(o) Babrius T(iti) f(ilius), C(aius) Capidas T(iti) f(ilius) C(ai) n(epos), V(ibius) Voisienus T(iti) f(ilius)
marones murum ab fornice ad circum et fornicem cisternamq(ue) d(e) s(enatus) s(ententia) faciundum coiravere.

@@

\begin{itemize}
\item for grammar: words of third declensions such as maro and fornix; complements of direction; gerundive to express
purpose; use of the form {}-ere in the perfect tense personal endings;
\item for history: the use of the Latin language prior to the Social War, as proof of the intense Romanization of the
area at the end of the II century B.C.\footnote{\citet{Coarelli1991}.}; the presence of elements in the names that are not
Roman but of Umbrian origin;
\item for archaeology: the building of terraces to create public spaces in a town like Assisi established on a hill; the
identification of the area around San Rufino as the ``acropolis''; the incorporation of the Roman wall into the left nave
of the church (the inscription being still in situ).
\end{itemize}

Schoolbooks rarely present these opportunities offered by inscriptions, so the main problem for a teacher is to access
appropriate material\footnote{ See the observations made by \citet{Carpenter2006}.}. Holding a PhD in Late Antiquity and
Umbrian inscriptions, I am fortunate enough to already know many sources. However, a graduate may not have enough
knowledge in epigraphy, may not be familiar with the Corpus Inscriptionum Latinarum; that is why the EAGLE project can
also offer tools for further insights.

@@

\begin{itemize}
\item \citet[39]{Barbieri2015}: the inscription of the architect C. Vettius Gratus (CIL X 3392) is
merely a decorative element on the page, without any relation to the topic (phonetic changes from Latin to Italian);

\item \citet[40]{Domenici2012}: programmata from Pompeii are introduced to explain the Roman
naming system;

\item \citet[269-270]{Gambis2013}: tabellae defixionum are used for
different purposes (demonstrative adjectives, functions of subjunctive); the subject (magic in the ancient world) may
not be relevant in a school context (the possession of such knowledge is not required), but sounds very intriguing to
students.
\end{itemize}

@@

Secondly, the style and the structures are different. Even when all abbreviations and integrations are explained (a
teacher should at least show the meaning of round and square brackets but should not ask a pupil to solve an
abbreviation, except the easy ones), the word order cannot immediately be reconstructed, especially in decrees and
carmina epigraphica. School dictionaries are not intended for interpreting epigraphical lexicon: some words may not be
present, their meaning may not always be explained. A lot of institutions and formulae which are clear to a specialist
(e.g. quattuorvir iure dicundo) may be hard to understand or to translate. It is, however, also the case that a teacher
cannot dedicate too much time to introducing these words to the class because his ultimate aim is different; so it is
better to focus on materials that do not contain much specific lexicon.

Examples:
\begin{itemize}
\item CIL 11, 04431 = EDR025160: \begin{quotation}
[Inf]austo, levis umbra, tuo mihi flebilis hora / sorte
tua certe tempus in omne fuit.

\end{quotation}
This funerary inscription from Ameria contains two verses, but the position of the words is tricky; infausto may be
taken as an adjective (as it is most commonly) and its meaning is basically the same as sorte tua; students who are not
used to poetry may not recognize the anastrophe tempus in omne.

 \item CIL 11, 04391 = EDR025123:\begin{quotation}
 Iuliae M(arci) f(iliae) Felicitati, / uxori C(ai) Curiati
Eutychetis / IIIIvir(i), magistrae Fortu/nae Mel(ioris), coll(egium) centonarior(um) / ob merita eius. Quo honore /
contenta sumptum omnem / remisit et ob dedic(ationem) ded(it) sin/gulis ((sestertios)) XX n(ummos) et hoc amplius /
arkae eorum intul(it) ((sestertium)) V m(ilia) n(ummum) / ut die natalis sui (ante diem) V Id(us) Mai(as) / ex usuris
eius summae epu/lantes imperpetuum divider(ent), / quod si divisio die s(upra) s(cripta) celebrata non / fuerit tunc
pertineb(it) omn(is) summa / ad familiam publicam.

\end{quotation}
Specialists are well used to an inscription like this. A high school student may find unexpected difficulties in
understanding the meaning of magistra (not teacher but priestess), but also centonarius or even hoc amplius; not to
mention the Roman calendar system, which always takes too much time to explain!

@@

The article published by F. Bigi in the Proceedings of the First EAGLE International Conference\footnote{\citet{Bigi2014}.}
includes many observations regarding the problems that may be encountered, for instance when translating names and
titles. In particular, I find the suggestion that round brackets should be used in the translations to provide further
explanations about specific offices rendered with the technical derivative word, or for concepts omitted in the
original Latin text, very useful. Examples of this could be words such as centonarii or marones seen in inscriptions
mentioned above, but also the following ones:

@@

\item CIL 11, 01925 = EDR142701: the names of the emperors M. Aurelius Antoninus and M.
Antoninus Pius Germaticus Sarmaticus need to be explained (Caracalla and Marcus Aurelius) to avoid confusion. It should
be noted that in my classes proper names, and not only those of emperors but also of other people, were generally
translated into Italian, even if it is advisable to transcribe them in the nominative case\footnote{
\url{http://www.eagle-network.eu/wiki/index.php/Guidelines\_for\_Translators}.}; the same thing was done with cognomina ex
virtute, considering that they are intuitively interpretable for an Italian.
\end{itemize}

@@

\section{Inscriptions and Literature: a few samples}
\noindent On a more advanced level, inscriptions may also integrate certain aspects related to the study of Latin Literature. The
most typical example could be a comparison between the Tabula Claudiana (CIL 13, 01668) and Tacitus’ account (Annales
XI, 23-24)\footnote{ On which see \citet{Jahn1993}.}: reading the original document is a privileged occasion to determine how
reliable the historian is when using his sources.

Highly original suggestions have been provided by Mauro Reali, who is also the author of different school manuals, in a
paper published online\footnote{\url{http://mediaclassica.loescher.it/nuove-e-\%93vecchie\%94-forme-di-multimedialita.n2799}
; see also \citet{Reali2015}.}. Being an expert on the subject, he offers a comparison between the ``noble'' form of
the political-philosophical amicitia presented in Cicero’s Laelius and the term \emph{amicus} mentioned in inscriptions
from the lower levels of society, such as CIL 05, 05300 from \emph{Comum} (a funerary stela made by a freedman for
Pliny the Younger) or CIL 05, 05923 = EDR124245 regarding a strange case of ``friend deletion'' long before the Facebook
era\footnote{ \url{http://www.laricerca.loescher.it/lingue-classiche/327-un-amico-o-amicus-e-per-sempre.html}.}.

@@

Viator, resiste et rogo / te et lege. Post annos XXVII ven[i] / Hispellum, in patriam meam. Scio / me oportere colere
hunc locum / ubi ossa meorum requiescunt et mea / et amicorum meorum. Ex hoc sepulch[ro] / cippi perierunt duo et
frontes duae sciun[t] / qui surupuit et acturi simus et legimus, / satis est testium etqs.

\end{quotation}
As for the Augustan age, Reali suggests showing some monumental inscriptions of the princeps (he impressively goes so
far as to compare the qualifications \emph{Imperator Caesar Augustus} to a modern logo or even a \emph{hashtag}). On
this subject, the altars Augusto sacrum put up by Perusia restituta (CIL 11, 01923 = EDR142666, EDR142667, EDR142668,
EDR142669), even with a simple text, offer the opportunity to deal with an aspect as crucial as the imperial cult.
Traditionally, we read that Augustus was worshipped directly only in the Eastern provinces, but not in Rome and Italy.
The inscriptions from Perusia testify that things are effectively more complex; these documents also offer remarkable
information about the restitutio of the town, destroyed at the end of the bellum in 40 B.C. A few years later Perusia
would become Augusta, as you can read on the city gates, especially on the newly restored Etruscan Arch (CIL 11, 01929
= EDR142706).

Teaching, as I do, in a school named after Propertius, I always stress the importance of reconstructing the origin of
the poet through the epigraphical data of the gens Propertia: the greatest number of written documents of the family
having been found in Assisi. For more than two hundred years, beginning with the Vois(ienus) Ner. (filius) Propertius
mentioned among the Umbrian magistrates of the late II century B.C., this gens stands out in the town for its social
influence and wealth\footnote{\citet{Forni1986}; \citet{Zuddas2006}.}. It is always very exciting to combine the information on
Passennus Paullus Propertius Blaesus given by CIL 11, 05405 = EDR025355 and that contained in two Letters by Pliny (VI,
15 and IX, 22). Pliny, showing great concern for his friend’s illness but also great esteem for him as an elegiac poet,
asserts that he is a descendant and fellow citizen of Propertius; the inscription (the front of an honorific base),
providing the full name, with the tribe Sergia typical of the inhabitants of Asisium, is indirect, but clear, evidence
that the Augustan poet was born there. The information provided by the inscription and the literary text integrates
perfectly. Students find it fascinating to look for traces of the poet inside the town, especially when they read the
following graffito on an interior wall of a Roman house underneath the church of Santa Maria Maggiore (EDR028769):

@@

[{}- - - I]ọvino consulibb(us) (ante diem) VIII Kal(endas) Martias domum oscilavi Musae. 
\end{quotation}

This house was still visited in the fourth century A.D., which really supports the theory that it used to be the poet’s
residence, and remained an object of reverence for centuries\footnote{\citet[244-246]{Boldrighini2014}.}. Through the stories
of Passennus Paullus and Sextus Propertius macro-history and local history meet to make the past come alive in every
corner of the modern town.

>>>>

The use of local materials, which can be checked out personally, may more easily arouse the learners’ attention. That is
why in this paper I will mainly refer to inscriptions from modern Umbria (regiones VI and VII, from Perusia) and
especially from Asisium, used at school during my experimental lessons. The local museum, best known as the ``Foro
romano'', hosts a huge collection of inscriptions that were catalogued in 2008 by pupils of the Liceo Properzio, under
my supervision, with details of the type of inscription, material, place of origin and chronology. Previously a booklet
about Roman Assisi had been
written,\footnote{\url{http://www.liceoassisi.it/index.php?option=com\_content\&view=article\&id=47:assisi-romana\&catid=6:i-nostri-risultati\&Itemid=34.}}
both in Italian and English, comprising a page about the famous tetrastyle with a translation of the inscription on the
base (CIL 11, 05372 = EDR025323, later inserted in the MediaWiki page), which recalls the official inauguration of the
aedicule: at the time, as was the Roman custom, money was given to decurions, seviri Augustales and the common people.
Thanks to these few lines, the class is taken on a journey to ancient Asisium, and gets a lot of feedback about
religion, architecture, social life and economy in the first century A.D.

@@

\begin{itemize}
\item elements of Latin grammar, syntax and linguistics;\footnote{\citet{Hartnett2012, McCarthy1992}.}
\item culture and history;
\item and also, to a lesser extent, lexicon. 
\end{itemize}

@@

\begin{itemize}
\item the ones that are relevant for their Italian derivates (e.g. \emph{os, oris});
\item the ones that belong to the most significant semantic fields (always keeping in mind that some words are more
important than others: for instance, knowing the difference between \emph{bellum} and \emph{pugna} helps to develop the comprehension
of the two different categories; on the contrary, a non-specialized student does not need to know who a \emph{primus pilus}
was);
\item the abstract ones that are fundamental to understanding the Roman way of thinking (e.g. imperium, virtus, fas).
\end{itemize}
Accordingly, a teacher should concentrate on those aspects of Roman society that are still pre-eminent for us:
archaeology (Roman buildings, domus, theatres, roads), myths and religion, politics (Empire, war and globalization).
Inscriptions certainly offer a less varied lexical repertoire than a literary text, but can still help to give
substance to these contexts.

@@

\begin{quotation}
\emph{P. Decimius P. l. Eros / Merula, medicus / clinicus, chirurgus, / ocularius, VIvir. / Hic pro libertate dedit
((sestertium)) (quinquaginta milia). / Hic pro seviratu in rem p(ublicam) / dedit ((sestertium)) (duo milia). / Hic in
statuas ponendas in / aedem Herculis dedit ((sestertium)) (triginta milia). / Hic in vias sternendas in / publicum
dedit ((sestertium)) (triginta septem milia). / Hic pridie quam mortuus est / reliquit patrimoni ((sestertium))... }
\end{quotation}

not only does a student make contact with the technical lexicon of medicine and words related to building activities
(\emph{statuas ponere, vias sternere}), but he can also perceive the ``evergetic spirit'' of an ancient society: a physician was
usually a freedman coming from the East, who could be very rich and spend his money for public utility; whenever a
person obtained a priesthood, he used to pay a \emph{summa honoraria}. The text can also be used for linguistic purposes
(prepositions like \emph{pro}; partitive genitive; deponent verbs).

This does not mean that inscriptions are not useful for seeing specific abstract words in their context: when talking
about Roman virtues and \emph{mos maiorum}, the famous \emph{Clipeus} from Arles (\emph{AE} 1952, 0165 = \emph{AE} 1994, 0227) is a
perfect complement to chapter 34 of the \emph{Res Gestae} (formally an inscription, too): \begin{quotation}
\emph{Senatus / populusque Romanus /
imp(eratori) Caesari / Divi f(ilio) / Augusto / co(n)s(uli) VIII dedit clupeum / virtutis, clementiae, / iustitiae,
pietatis erga / deos patriamque.}

@@

For instance, if I say\emph{ Iulia rosas amat}, even an Italian who knows no Latin can get the meaning of the sentence.
Problems are encountered when the sentences and texts to translate become more complex: only when it is too late does
an ill-prepared teacher realize that the class has not internalized the language system and the patterns.

In funerary inscriptions the value of cases is essential. The texts can be simple, but if you want to understand who is
dead, who is the dedicant and what the relationship between them is, you have to distinguish dative and nominative
cases. Moreover, the concordance of the different elements of the onomastics (especially \emph{praenomina} and filiation) is
good way to encourage (and, for a teacher, to test) the learning of declensions not in a merely mnemonic, but also in
an active way.

@@

\begin{itemize}
\item CIL 11, 05501 = EDR025449: \begin{quotation}
\emph{Noniae Privatae, C. Propertius -{}-{}-{}-{}-{}-}
\end{quotation}

@@

\item CIL 11, 05461 = EDR025411: \begin{quotation}
\emph{A(n)noru(m) XIX. Calventia C(ai) f(ilia) Polla, L(ucius)
Vistinius vir, Gavia mater posuer(unt). }

@@

\item CIL 11, 05399 = EDR025349: \begin{quotation}
\emph{P(ublius) Decimius P(ubli) l(ibertus) Eros Merula VIvir
viam a cisterna ad domum L(uci) Muti stravit ea pecunia -{}-{}-{}-{}-{}- }
\end{quotation}

With an inscription like this you can address different topics, both historical (the importance of freedmen, the
imperial cult) and linguistic ones (indirect complements, uses of \emph{is ea id} as an adjective and to introduce a relative
clause that, in this case, has evidently been lost).
\end{itemize}

@@

\begin{quotation}
\emph{Post(umus) Mimesius C(ai) f(ilius), T(itus) Mimesius Sert(oris) f(ilius), Ner(o) Capidas C(ai) f(ilius) Ruf(- - -),
Ner(o) Babrius T(iti) f(ilius), C(aius) Capidas T(iti) f(ilius) C(ai) n(epos), V(ibius) Voisienus T(iti) f(ilius)
marones murum ab fornice ad circum et fornicem cisternamq(ue) d(e) s(enatus) s(ententia) faciundum coiravere.}

@@

\begin{itemize}
\item for grammar: words of third declensions such as \emph{maro} and \emph{fornix}; complements of direction; gerundive to express
purpose; use of the form {}\emph{-ere} in the perfect tense personal endings;
\item for history: the use of the Latin language prior to the Social War, as proof of the intense Romanization of the
area at the end of the II century B.C.;\footnote{\citet{Coarelli1991}.} the presence of elements in the names that are not
Roman but of Umbrian origin;
\item for archaeology: the building of terraces to create public spaces in a town like Assisi established on a hill; the
identification of the area around San Rufino as the ``acropolis''; the incorporation of the Roman wall into the left nave
of the church (the inscription being still in situ).
\end{itemize}

Schoolbooks rarely present these opportunities offered by inscriptions, so the main problem for a teacher is to access
appropriate material.\footnote{ See the observations made by \citet{Carpenter2006}.} Holding a PhD in Late Antiquity and
Umbrian inscriptions, I am fortunate enough to already know many sources. However, a graduate may not have enough
knowledge in epigraphy, may not be familiar with the \emph{Corpus Inscriptionum Latinarum}; that is why the EAGLE project can
also offer tools for further insights.

@@

\begin{itemize}
\item \citet[39]{Barbieri2015}: the inscription of the architect \emph{C. Vettius Gratus} (CIL X 3392) is
merely a decorative element on the page, without any relation to the topic (phonetic changes from Latin to Italian);

\item \citet[40]{Domenici2012}: \emph{programmata} from Pompeii are introduced to explain the Roman
naming system;

\item \citet[269-270]{Gambis2013}: \emph{tabellae defixionum} are used for
different purposes (demonstrative adjectives, functions of subjunctive); the subject (magic in the ancient world) may
not be relevant in a school context (the possession of such knowledge is not required), but sounds very intriguing to
students.
\end{itemize}

@@

Secondly, the style and the structures are different. Even when all abbreviations and integrations are explained (a
teacher should at least show the meaning of round and square brackets but should not ask a pupil to solve an
abbreviation, except the easy ones), the word order cannot immediately be reconstructed, especially in decrees and
carmina epigraphica. School dictionaries are not intended for interpreting epigraphical lexicon: some words may not be
present, their meaning may not always be explained. A lot of institutions and \emph{formulae} which are clear to a specialist
(e.g. \emph{quattuorvir iure dicundo}) may be hard to understand or to translate. It is, however, also the case that a teacher
cannot dedicate too much time to introducing these words to the class because his ultimate aim is different; so it is
better to focus on materials that do not contain much specific lexicon.

Examples:
\begin{itemize}
\item CIL 11, 04431 = EDR025160: \begin{quotation}
\emph{[Inf]austo, levis umbra, tuo mihi flebilis hora / sorte
tua certe tempus in omne fuit.}

\end{quotation}
This funerary inscription from Ameria contains two verses, but the position of the words is tricky; \emph{infausto} may be
taken as an adjective (as it is most commonly) and its meaning is basically the same as \emph{sorte tua}; students who are not
used to poetry may not recognize the anastrophe \emph{tempus in omne}.

 \item CIL 11, 04391 = EDR025123:\begin{quotation}
\emph{ Iuliae M(arci) f(iliae) Felicitati, / uxori C(ai) Curiati
Eutychetis / IIIIvir(i), magistrae Fortu/nae Mel(ioris), coll(egium) centonarior(um) / ob merita eius. Quo honore /
contenta sumptum omnem / remisit et ob dedic(ationem) ded(it) sin/gulis ((sestertios)) XX n(ummos) et hoc amplius /
arkae eorum intul(it) ((sestertium)) V m(ilia) n(ummum) / ut die natalis sui (ante diem) V Id(us) Mai(as) / ex usuris
eius summae epu/lantes imperpetuum divider(ent), / quod si divisio die s(upra) s(cripta) celebrata non / fuerit tunc
pertineb(it) omn(is) summa / ad familiam publicam. }

\end{quotation}
Specialists are well used to an inscription like this. A high school student may find unexpected difficulties in
understanding the meaning of \emph{magistra} (not teacher but priestess), but also \emph{centonarius} or even \emph{hoc amplius}; not to
mention the Roman calendar system, which always takes too much time to explain!

@@

The article published by F. Bigi in the Proceedings of the First EAGLE International Conference\footnote{\citet{Bigi2014}.}
includes many observations regarding the problems that may be encountered, for instance when translating names and
titles. In particular, I find the suggestion that round brackets should be used in the translations to provide further
explanations about specific offices rendered with the technical derivative word, or for concepts omitted in the
original Latin text, very useful. Examples of this could be words such as \emph{centonarii} or \emph{marones} seen in inscriptions
mentioned above, but also the following ones:

@@

\item CIL 11, 01925 = EDR142701: the names of the emperors \emph{M. Aurelius Antoninus} and \emph{M.
Antoninus Pius Germaticus Sarmaticus} need to be explained (Caracalla and Marcus Aurelius) to avoid confusion. It should
be noted that in my classes proper names, and not only those of emperors but also of other people, were generally
translated into Italian, even if it is advisable to transcribe them in the nominative case;\footnote{
\url{http://www.eagle-network.eu/wiki/index.php/Guidelines\_for\_Translators}.} the same thing was done with \emph{cognomina ex
virtute}, considering that they are intuitively interpretable for an Italian.
\end{itemize}

@@

\section{Inscriptions and Literature: a few samples}
\noindent On a more advanced level, inscriptions may also integrate certain aspects related to the study of Latin Literature. The
most typical example could be a comparison between the Tabula Claudiana (CIL 13, 01668) and Tacitus’ account (Annales
XI, 23-24):\footnote{On which see \citet{Jahn1993}.} reading the original document is a privileged occasion to determine how
reliable the historian is when using his sources.

Highly original suggestions have been provided by Mauro Reali, who is also the author of different school manuals, in a
paper published online.\footnote{\url{http://mediaclassica.loescher.it/nuove-e-\%93vecchie\%94-forme-di-multimedialita.n2799}
; see also \citet{Reali2015}.} Being an expert on the subject, he offers a comparison between the ``noble'' form of
the political-philosophical \emph{amicitia} presented in Cicero’s \emph{Laelius} and the term \emph{amicus} mentioned in inscriptions
from the lower levels of society, such as CIL 05, 05300 from \emph{Comum} (a funerary stela made by a freedman for
Pliny the Younger) or CIL 05, 05923 = EDR124245 regarding a strange case of ``friend deletion'' long before the Facebook
era.\footnote{ \url{http://www.laricerca.loescher.it/lingue-classiche/327-un-amico-o-amicus-e-per-sempre.html}.}

@@

\emph{Viator, resiste et rogo / te et lege. Post annos XXVII ven[i] / Hispellum, in patriam meam. Scio / me oportere colere
hunc locum / ubi ossa meorum requiescunt et mea / et amicorum meorum. Ex hoc sepulch[ro] / cippi perierunt duo et
frontes duae sciun[t] / qui surupuit et acturi simus et legimus, / satis est testium etqs.}

\end{quotation}
As for the Augustan age, Reali suggests showing some monumental inscriptions of the \emph{princeps} (he impressively goes so
far as to compare the qualifications \emph{Imperator Caesar Augustus} to a modern logo or even a \emph{hashtag}). On
this subject, the altars \emph{Augusto sacrum} put up by \emph{Perusia restituta} (CIL 11, 01923 = EDR142666, EDR142667, EDR142668,
EDR142669), even with a simple text, offer the opportunity to deal with an aspect as crucial as the imperial cult.
Traditionally, we read that Augustus was worshipped directly only in the Eastern provinces, but not in Rome and Italy.
The inscriptions from \emph{Perusia} testify that things are effectively more complex; these documents also offer remarkable
information about the \emph{restitutio} of the town, destroyed at the end of the \emph{bellum} in 40 B.C. A few years later \emph{Perusia}
would become \emph{Augusta}, as you can read on the city gates, especially on the newly restored Etruscan Arch (CIL 11, 01929
= EDR142706).

Teaching, as I do, in a school named after Propertius, I always stress the importance of reconstructing the origin of
the poet through the epigraphical data of the \emph{gens Propertia}: the greatest number of written documents of the family
having been found in Assisi. For more than two hundred years, beginning with the \emph{Vois(ienus) Ner. (filius) Propertius}
mentioned among the Umbrian magistrates of the late II century B.C., this gens stands out in the town for its social
influence and wealth.\footnote{\citet{Forni1986}; \citet{Zuddas2006}.} It is always very exciting to combine the information on
\emph{Passennus Paullus Propertius Blaesus }given by CIL 11, 05405 = EDR025355 and that contained in two Letters by Pliny (VI,
15 and IX, 22). Pliny, showing great concern for his friend’s illness but also great esteem for him as an elegiac poet,
asserts that he is a descendant and fellow citizen of Propertius; the inscription (the front of an honorific base),
providing the full name, with the tribe Sergia typical of the inhabitants of \emph{Asisium}, is indirect, but clear, evidence
that the Augustan poet was born there. The information provided by the inscription and the literary text integrates
perfectly. Students find it fascinating to look for traces of the poet inside the town, especially when they read the
following graffito on an interior wall of a Roman house underneath the church of Santa Maria Maggiore (EDR028769):

@@

\emph{[{}- - - I]ọvino consulibb(us) (ante diem) VIII Kal(endas) Martias domum oscilavi Musae.}
\end{quotation}

This house was still visited in the fourth century A.D., which really supports the theory that it used to be the poet’s
residence, and remained an object of reverence for centuries.\footnote{\citet[244-246]{Boldrighini2014}.} Through the stories
of Passennus Paullus and Sextus Propertius macro-history and local history meet to make the past come alive in every
corner of the modern town.

### 2016-01-20 source/3/EAGLE_2016_submission_3.tex
fixed few more papers
minor changes. THERE HAS BEEN CONFUSION WITH THE BIBLIO, MASTER AND
SOME OTHER FILES

The database in itself would have three components (see below Fig.~\ref{fig:schemadatabase}), following the best practices in the field
(\citet{doi:10.3200/HMTS.37.1.34-38}): the \emph{sources database} (with ``facsimile'' transcription of the sources’ text), the \emph{central database} (the complete, corrected, integrated, standardized and coded form of the sources database) and the \emph{data releases} (destined for on-line usage, allowing easy extraction of data in view of analyses).

@@

Regarding the data entry, we are trying to remain faithful to the source and to record, during the first phase, only the
minimum of deduced information (gender, juridical status, ethnicity of the name). At the same time, we operate some
conventional onomastic transcriptions (AEL will be Aelius from the start, etc.). All these basic normalization
procedures are being thoroughly explained and documented in the data entry manual, and the reason why we chose to apply
them is to speed up data standardization, run analyses and publish results on small samples, whose standardization
procedures would otherwise be overwhelmingly time-consuming in relation with the results.

@@

The first table to be filled in is the file of the source – \emph{Inscriptions}. First of all, each inscription gets a source
code, formed of 5 digits and a symbol/acronym of the province’s name (D for Dacia, MS for Moesia Superior, DAL for
Dalmatia, etc.) – so we have, for example 00001MS. The \emph{Inscriptions} table has text fields, as well as value lists.
Then, we have text strings for: \emph{Relevant expressions}, \emph{Stylistic details}, \emph{Atypical features}, \emph{Observations}, \emph{Place of discovery}, \emph{Place of provenience}, \emph{Ancient name provenience}, \emph{Timestamp/Timeframe} and \emph{External links}. Although we are
aware that some of the data (\emph{Timestamp/Timeframe}) could have benefitted from a standardized form, at this point we
opted for vaster possibilities of expression and adaptation. Other information will be filled using value lists, as
standardization is more suited: \emph{Type of inscription}, \emph{Language}, \emph{Material}. For these we will use the Eagle
vocabularies,\footnote{\url{http://www.eagle-network.eu/resources/vocabularies}} because they offer an already
standardized language.

>>>>

The database in itself would have three components (see below Fig.~\ref{fig:schemadatabase}), following the best practices in the field
\citep{doi:10.3200/HMTS.37.1.34-38}: the \emph{sources database} (with ``facsimile'' transcription of the sources’ text), the \emph{central database} (the complete, corrected, integrated, standardized and coded form of the sources database) and the \emph{data releases} (destined for on-line usage, allowing easy extraction of data in view of analyses).

@@

Regarding the data entry, we are trying to remain faithful to the source and to record, during the first phase, only the
minimum of deduced information (gender, juridical status, ethnicity of the name). At the same time, we operate some
conventional onomastic transcriptions (AEL will be \emph{Aelius} from the start, etc.). All these basic normalization
procedures are being thoroughly explained and documented in the data entry manual, and the reason why we chose to apply
them is to speed up data standardization, run analyses and publish results on small samples, whose standardization
procedures would otherwise be overwhelmingly time-consuming in relation with the results.

@@

The first table to be filled in is the file of the source – \emph{Inscriptions}. First of all, each inscription gets a source
code, formed of 5 digits and a symbol/acronym of the province’s name (D for \emph{Dacia}, MS for \emph{Moesia Superior}, DAL for
\emph{Dalmatia}, etc.) – so we have, for example 00001MS. The \emph{Inscriptions} table has text fields, as well as value lists.
Then, we have text strings for: \emph{Relevant expressions}, \emph{Stylistic details}, \emph{Atypical features}, \emph{Observations}, \emph{Place of discovery}, \emph{Place of provenience}, \emph{Ancient name provenience}, \emph{Timestamp/Timeframe} and \emph{External links}. Although we are
aware that some of the data (\emph{Timestamp/Timeframe}) could have benefitted from a standardized form, at this point we
opted for vaster possibilities of expression and adaptation. Other information will be filled using value lists, as
standardization is more suited: \emph{Type of inscription}, \emph{Language}, \emph{Material}. For these we will use the Eagle
vocabularies,\footnote{\url{http://www.eagle-network.eu/resources/vocabularies}} because they offer an already
standardized language.

### 2016-01-20 source/36/DatabasesToEpiDoc.tex
fixed few more papers
minor changes. THERE HAS BEEN CONFUSION WITH THE BIBLIO, MASTER AND
SOME OTHER FILES

There are several mappings involved in the aggregation work of EAGLE. Content providers need to map to a common Eagle Metadata Format, and the data produced will then be mapped to the Europeana Data Model\footnote{\citet{D3.1}} in order to be aggregated in Europeana, for a wider dissemination with a special eye for the general user.

@@

Given the template described in Part III, and ANNEX II of the EAGLE Metadata Schema (\citet{D3.1}), an XSL Transformation converts from string epigraphic texts in marked up TEI-EPIDOC XML, following the \href{http://www.stoa.org/epidoc/gl/dev/}{EpiDoc guidelines} (\citet{Elliott2007}).

>>>>

There are several mappings involved in the aggregation work of EAGLE. Content providers need to map to a common Eagle Metadata Format, and the data produced will then be mapped to the Europeana Data Model\footnote{\citet{sicilia_eagle_2015}} in order to be aggregated in Europeana, for a wider dissemination with a special eye for the general user.

@@

Given the template described in Part III, and ANNEX II of the EAGLE Metadata Schema (\citet{sicilia_eagle_2015}), an XSL Transformation converts from string epigraphic texts in marked up TEI-EPIDOC XML, following the \href{http://www.stoa.org/epidoc/gl/dev/}{EpiDoc guidelines} (\citet{Elliott2007}).
